Temporal stability in human interaction networks: sector sizes, topological prominence and time activity

### 2015-10-04 paper.tex
commit intermediario p nao perder ao arrumar equacao

\begin{table}
	\centering
	\caption{Columns $date_1$ and $date_M$ have dates of first and last messages from the 20,000 messages considered in each email list.
		$N$ is the number of participants (number of different email addresses).
		$\Gamma$ is the number of discussion threads (count of messages without antecedent).
		$\overline{M}$ is the number of messages missing in the 20,000 collection, $100\frac{23}{20000}=0.115$ percent in the worst case.
	}
	\label{tab:genLists}
	\begin{tabular}{|l|c|c|c|c|c|}\hline
		list & $date_1$ & $date_{M}$    & $N$  & $\Gamma$ & $\overline{M}$ \\\hline
		\input{tables/tab1Geral}
	\end{tabular}
\end{table}

@@

\section{Methods}\label{sec:carac}
The email lists and the networks generated from them were characterized using four procedures, namely: 1) statistics of activity along time, in scales from seconds to years; 2) sectioning of the networks in hubs, intermediary and peripheral vertices; 3) dispersion of basic topological metrics; 4) iterative visualization and data inspection.
These procedures are described below.

\subsection{Time activity statistics}\label{sec:mtime}
Messages were counted over time with respect to seconds, minutes, hours, days of the week, days of the month, and months of the year. This resulted in histograms from which patterns could be drawn. The ratio $\frac{b_h}{b_l}$ between the highest and lowest incidences on the histograms served as a clue of how close the observed distribution is to a uniform distribution.

Regular measures of location and dispersion, such as usual mean and
standard deviation, hold little meaning when observations take
place in a compact Riemannian manifold.
Therefore, such measures were taken using circular statistics, in which each $measurement$ (data point) is represented as a unity complex number, $z=e^{i\theta}=\cos(\theta)+i\sin(\theta)$, where $\theta=measurement\frac{2\pi}{T}$, T is the period in which the counting is repeated. For example, $\theta=12\frac{2\pi}{24}=\pi$ for a message sent at $12h$ and given $T=24h$ for days. The moments $m_n$, lengths of moments $R_n$, mean angle $\theta_\mu$, and rescaled mean angle $\theta_\mu'$ are defined as:

@@

\noindent As expected, a positive correlation was found in all $Var(z)$, $S(z)$ and $\delta(z)$ dispersion measures, as can be noticed in Section~\ref{si:circ} of the Supporting Information, and $\delta(z)$ was preferred in the discussion of results.

\subsection{Interaction networks}\label{intNet}
Interaction networks can be modeled both weighted or unweighted, both directed or undirected~\cite{bird,newmanCommunityDirected,newmanCommunity2013}.
Networks in this paper are directed and weighted, the most informative of trivial possibilities. We did not investigate directed unweighted, undirected weighted, and undirected unweighted representations of the interaction networks. 
The networks were obtained as follows: a direct response from participant B to a message from participant A yields an edge from A to B, as information went from A to B. The reasoning is: if B wrote a response to a message from A, he/she read what A wrote and formulated a response, so B assimilated information from A, thus $A \rightarrow B$. Inverting edge direction yields the status network: B read the message and considered what A wrote worth responding, giving status to A, thus $B\rightarrow A$. This paper uses the information network as described above and depicted in Figure~\ref{formationNetwork}. Edges in both directions are allowed. Each time an interaction occurs, the value of one is added to the edge weight. Selfloops were regarded as non-informative and discarded. These human social interaction networks are reported in the literature as exhibiting scale-free and small world properties, as expected for (some) social networks~\cite{bird,newmanBook}.

\begin{figure}[!h]
	\centering
	\includegraphics[width=0.44\textwidth]{figs/criaRede2_}
	\caption{The formation of interaction networks from email messages. Each vertex represents a participant. A reply message from author B to a message from author A is regarded as evidence that B received information from A and yields a directed edge. Multiple messages add ``weight'' to a directed edge. Further details are given in Section~\ref{intNet}.}
	\label{formationNetwork}
\end{figure}

@@

%\subsubsection{Sectioning of networks in peripheral, intermediary and hubs sectors}\label{sectioning}
\subsection{Principal Component Analysis of topological metrics}\label{sec:pca}
Principal Component Analysis (PCA) is a well documented technique~\cite{pca} and was used to learn: 1) which metrics contribute to each principal component and in which proportion; 2) how much of the dispersion is concentrated in each component; 3) expected values and dispersions for these quantities over various networks.

Let $\mathbf{X}=\{X[i,j]\}$ be a matrix of all vertices $i$ and the respective
values for each metric $j$, 
$\mu_X [j]=\frac{\sum_i X[i,j]}{I}$ the mean of metric $j$ over all $I$ vertices, 
$\sigma_X [j]=\sqrt{\frac{\sum_i (X[i,j]-\mu_X [j])^2}{I}}$ the standard deviation of metric $j$,
and $\mathbf{X'}=X'[i,j]={\frac{X[i,j]-\mu_X[j]}{\sigma_X[j]}}$ 
the matrix with the \emph{z-score} of each metric. 
Let $\mathbf{V}=\{V[j,k]\}$ be the matrix $J\times J$ of eigenvectors
of the covariance matrix $\mathbf{C}$
of $\mathbf{X'}$, one eigenvector per column.
Each eigenvector combines the original measures into one principal component, therefore
$V'[j,k]=100*\frac{|V[j,k]|}{\sum_{j'} |V[j',k]|}$
delivers the percentage of the principal component $k$
that is due to the measure $j$.
With $k$ eigenvectors 
$D[k]$,
it is enough to observe $D'[k]=100*\frac{D[k]}{\sum_{k'}D[k']}$
to know the percentage of dispersion that each principal component
is responsible for.
With the eigenvalues $k$ in ascending order
we observe, in general, the three first eigenvalues and
the respective eigenvectors in percentages:
$\{(V'[j,k],\;D'[k])\}$.
These usually reveal sufficient patterns for a sound analysis and sum between 60 and 95\% of the dispersion.
Namely, given $L$ observations $l$, each with $k$ pairs of
eigenvalues and eigenvectors, we observe
the mean
$\mu_{V'}[j,k]$
and the standard deviation $\sigma_{V'}[j,k]$ 
of the metric $j$ in the principal component $k$,
and the mean
$\mu_{D'}[k]$
and the standard deviation 
$\sigma_{D'}[k]$
of the contribution of the component $k$ to the dispersion
of the system:

The covariance matrix 
$\mathbf{C}$ is the correlation matrix because $\mathbf{X'}$ is normalized. Therefore, it is also directly observed as a first clue for patterns.
This was accomplished with the most simple associations:
low absolute values indicate low correlation (and a possible independence);
high values indicate positive correlation; negative values with high absolute value indicate negative correlation.

\subsubsection{Metrics considered}\label{measures}

The topology of the networks was studied using PCA with a small selection of the most basic and fundamental measurements for each vertex, as follows:

@@

@@

The lower degree fat tail corresponds to the border vertices, i.e. the peripheral sector or periphery where $\widetilde{P}(k)>P(k)$ and $k$ is lower than any intermediary sector value of $k$. The higher degree fat tail is the hub sector, i.e. $\widetilde{P}(k)>P(k)$ and $k$ is higher than any intermediary sector value of $k$. The reasoning for this classification is as follows: vertices so connected that they are virtually inexistent in networks connected at pure chance (e.g. without preferential attachment) are correctly associated to the hubs sector. Vertices with very few connections, which are way more abundant than expected by pure chance, are assigned to the periphery. Vertices with degree values predicted as the most abundant if connections are created by pure chance, near the average, and less frequent in the real network, are classified as intermediary.

\begin{figure}[!h]
	\centering
	\includegraphics[width=0.5\textwidth]{figs/fser__}
	\caption{Classification of vertices by comparing the degree
		distributions.
		The binomial distribution of the Erd\"os R\'enyi network exhibit more intermediary vertices, while a scale-free network, associated with the power-law distribution,
		has more peripheral and hub vertices.
		The sector borders are defined with respect to the intersections of the distributions.
Characteristic degrees are in the compact intervals: $[0,k_L]$, $(k_L,k_R]$, $(k_R,k_{max}]$ for the periphery, intermediary and hubs sectors (the ``Erd\"os sectors'').
		The empirical distributions of connectivity in interaction networks, e.g. derived from email lists,
		can be sectioned by comparison with to the associated binomial distribution.
		In this figure, a snapshot of 1000 messages from CPP list yield the degree distribution
		of an interaction network of 98 nodes and 235 edges.
		A throughout exposition of the method is exposed in Section~\ref{sectioning}.}
		\label{fig:setores}
\end{figure}
To ensure statistical validity of the histograms, bins can be chosen to contain at least $\eta$ vertices of the real network. The range $\Delta$ of incident values should be partitioned in $m$ parts $\Delta=\cup_{i=1}^m \Delta_i$, with $\Delta_i\cap \Delta_j \; \forall\; i \neq j$. Thus, $\Delta_i=\{max(\Delta_{i-1})<k\leq j \;\;|\; \sum_{max(\Delta_{i-1})+1}^j \eta_k \geq \eta \text{ and } (\sum_{max(\Delta_{i-1})+1}^{j-1} \eta_k < \eta \text{ or } j=max(\Delta_{i-1})+1) \}$, with $\eta_k$ the number of vertices with degree $k$ and $max(\Delta_{0})=-1$. This changes equation~\ref{criterio} to

>>>>

\begin{table}
	\centering
	\caption{Columns $date_1$ and $date_M$ have dates of first and last messages from the 20,000 messages considered in each email list.
		$N$ is the number of participants (number of different email addresses),
		$\Gamma$ is the number of discussion threads (count of messages without antecedent),
		$\overline{M}$ is the number of messages missing in the 20,000 collection
		($100\frac{23}{20000}=0.115$ percent in the worst case).
	}
	\label{tab:genLists}
	\begin{tabular}{|l|c|c|c|c|c|}\hline
		list & $date_1$ & $date_{M}$    & $N$  & $\Gamma$ & $\overline{M}$ \\\hline
		\input{tables/tab1Geral}
	\end{tabular}
\end{table}

@@

\section{Methods}\label{sec:carac}
%The networks were characterized with: 1) statistics of activity along time, in scales from seconds to years; 2) dispersion of basic topological metrics; 3) sectioning of the networks in hubs, intermediary and periphery; 4) iterative visualization, sonification and data inspection.
%These procedures are described below.

\subsection{Time activity statistics}\label{sec:mtime}
Messages were counted over time as histograms in terms of seconds,
minutes, hours, days of the week, days of the month, and months of the year.
Most standard measures of location and dispersion, such as usual mean and
standard deviation, hold little meaning in a compact Riemannian manifold.
Equivalent measures were taken using circular statistics,
in which each measurement $t$ (data point) is represented as a unity complex number,
$z=e^{i\theta}=\cos(\theta)+i\sin(\theta)$, where $\theta=t\frac{2\pi}{T}$,
and $T$ is the period in which the counting is repeated.
For example, $\theta=12\frac{2\pi}{24}=\pi$ for a message sent at $t=12h$ and given $T=24h$ for days.
The moments $m_n$, lengths of moments $R_n$, mean angles $\theta_\mu$, and rescaled mean angles $\theta_\mu'$ are defined as:

@@

\noindent
Also, the ratio $r=\frac{b_l}{b_h}$ between the lowest and the highest incidences on the histograms 
served as a further clue of how close the distribution was to being uniform.
As expected, a positive correlation was found in all $r, Var(z)$, $S(z)$ and $\delta(z)$ dispersion measures,
which can be noticed in the Section~\ref{si:circ} of the Supporting Information document.
The circular dispersion $\delta(z)$ was found more sensitive and therefore preferred in the discussion of results.

\subsection{Interaction networks}\label{intNet}
Interaction networks can be modeled both weighted or unweighted,
both directed or undirected~\cite{bird,newmanCommunityDirected,newmanCommunity2013}.
Networks in this paper are directed and weighted, the most informative of trivial possibilities.
We did not investigate directed unweighted, undirected weighted,
and undirected unweighted representations of the interaction networks.

The interaction networks were obtained as follows:
a direct response from participant B to a message from participant A yields an edge from A to B, as information went from A to B.
The reasoning is: if B wrote a response to a message from A,
he/she read what A wrote and formulated a response, so B assimilated information from A, thus $A \rightarrow B$.
Edges in both directions are allowed.
Each time an interaction occurs, the value of one is added to the edge weight.
Selfloops were regarded as non-informative and discarded.
Inverting edge direction yields the status network:
B read the message and considered what A wrote worth responding, giving status to A,
thus $B\rightarrow A$.
This paper considers by convention the information network as described above ($A\rightarrow B$) and depicted in Figure~\ref{formationNetwork}.
These human social information interaction networks are reported in the literature
as exhibiting scale-free and small-world properties,
as expected for a number of social networks~\cite{bird,newmanBook}.

\begin{figure}[!h]
	\centering
	\includegraphics[width=0.44\textwidth]{figs/criaRede2_}
	\caption{The formation of interaction networks from exchanged messages.
	Each vertex represents a participant.
	A reply message from author B to a message from author A is regarded as evidence
	that B received information from A and yields a directed edge.
	Multiple messages add ``weight'' to a directed edge.
	Further details are given in Section~\ref{intNet}.}
	\label{formationNetwork}
\end{figure}

@@

\subsubsection{Topological metrics}\label{measures}

The topology of the networks was studied using a small selection of the most basic and fundamental measurements for each vertex, as follows:

@@

%\subsubsection{Sectioning of networks in peripheral, intermediary and hubs sectors}\label{sectioning}
\subsection{Principal Component Analysis of topological metrics}\label{sec:pca}
Principal Component Analysis (PCA) is a well documented technique~\cite{pca} and was used to acquire knowledge about:
1) which metrics contribute to each principal component and in which proportion; 2) how much of the dispersion is concentrated in each component; 3) expected values and dispersions for these quantities over various networks.

Let $\mathbf{X}=\{X[i,j]\}$ be a matrix of all vertices $i$ and the respective
values for each metric $j$, 
$\mu_X [j]=\frac{\sum_i X[i,j]}{I}$ the mean of metric $j$ over all $I$ vertices, 
$\sigma_X [j]=\sqrt{\frac{\sum_i (X[i,j]-\mu_X [j])^2}{I}}$ the standard deviation of metric $j$,
and $\mathbf{X'}=\{X'[i,j]\}=\left\{\frac{X[i,j]-\mu_X[j]}{\sigma_X[j]}\right\}$ 
the matrix with the \emph{z-score} of each metric. 
Let $\mathbf{V}=\{V[j,k]\}$ be the matrix $J\times J$ of eigenvectors
of the covariance matrix $\mathbf{C}$
of $\mathbf{X'}$, one eigenvector per column.
Each eigenvector combines the original measures into one principal component, therefore
$V'[j,k]=100\frac{|V[j,k]|}{\sum_{j'} |V[j',k]|}$
is the percentage of the principal component $k$
that is directly proportional to the measure $j$.
%With $k$ eigenvectors 
%$D[k]$,
%it is enough to 
Be $\mathbf{D}=\{D[k]\}$ the eigenvalues associated to the eigenvectors $\mathbf{V}$,
then $D'[k]=100*\frac{D[k]}{\sum_{k'}D[k']}$
is the percentage of total dispersion of the system that the principal component $k$
is responsible for.
We consider, in general, the three greatest eigenvalues and
the respective eigenvectors in percentages:
$\{(D'[k],\;V'[j,k])\}$.
These usually sum between 60 and 95\% of the dispersion
and reveal patterns for a sound analysis.
In particular, 
given $L$ snapshots $l$ of the interaction network,
we are interested in the mean
$\mu_{V'}[j,k]$
and the standard deviation $\sigma_{V'}[j,k]$ 
of the contribution of metric $j$ to the principal component $k$,
and the mean
$\mu_{D'}[k]$
and the standard deviation 
$\sigma_{D'}[k]$
of the contribution of the component $k$ to the dispersion
of the system:

The covariance matrix 
$\mathbf{C}$ is the correlation matrix because $\mathbf{X'}$ is normalized.
Therefore, $\mathbf{C}$ is also directly observed as a first clue for patterns
by the most simple associations:
low absolute values indicate low correlation (and a possible independence);
high values indicate positive correlation;
negative values with a high absolute value indicate negative correlation.

@@

The lower degree fat tail corresponds to the border vertices, i.e. the peripheral sector or periphery where $\widetilde{P}(k)>P(k)$ and $k$ is lower than any intermediary sector value of $k$. The higher degree fat tail is the hub sector, i.e. $\widetilde{P}(k)>P(k)$ and $k$ is higher than any intermediary sector value of $k$. The reasoning for this classification is as follows: vertices so connected that they are virtually inexistent in networks connected at pure chance (e.g. without preferential attachment and as in the Erd\"os R\'enyi model) are correctly associated to the hubs sector. Vertices with very few connections, which are way more abundant than expected by pure chance, are assigned to the periphery. Vertices with degree values predicted as the most abundant if connections are created by pure chance, near the average, and less frequent in the real network, are classified as intermediary.

\begin{figure}[!h]
	\centering
	\includegraphics[width=0.5\textwidth]{figs/fser__}
	\caption{Classification of vertices by comparing degree
		distributions.
		The binomial distribution of the Erd\"os R\'enyi network exhibit more intermediary vertices, while a scale-free network, associated with the power-law distribution,
		has more peripheral and hub vertices.
		The sector borders are defined with respect to the intersections of the distributions.
Characteristic degrees are in the compact intervals: $[0,k_L]$, $(k_L,k_R]$, $(k_R,k_{max}]$ for the periphery,
intermediary and hubs sectors, the ``Erd\"os sectors''.
		The connectivity distribution of empirical interaction networks, e.g. derived from email lists,
		can be sectioned by comparison against the associated binomial distribution with the same number of vertices and edges.
		In this figure, a snapshot of 1000 messages from CPP list yield the degree distribution
		of an interaction network of 98 nodes and 235 edges.
		A throughout exposition of the method is exposed in Section~\ref{sectioning}.}
		\label{fig:setores}
\end{figure}
To ensure statistical validity of the histograms, bins can be chosen to contain at least $\eta$ vertices of the real network.
The range $\Delta$ of incident values should be partitioned in $m$ parts $\Delta=\cup_{i=1}^m \Delta_i$,
with $\Delta_i\cap \Delta_j=\emptyset \; \forall\; i \neq j$:
%\begin{equation}
%\begin{split}
\begin{align}
\Delta_i =\left\{ k & | & \overline{\Delta}_{i-1}<k\leq j \text{ and }\\
& [ & N - \sum_{k=0}^{\overline{\Delta}_{i-1}}< \eta \text{ or }\\
&	[ & \sum_{k=\overline{\Delta}_{i-1}+1}^j \eta_k \geq \eta \text{ and }\\
&(&\sum_{\overline{\Delta}_{i-1}+1}^{j-1} \eta_k < \eta \text{ or }   j=\overline{\Delta}_{i-1}+1)] ] \rignt\}
\end{align}
%\end{split}
%\end{equation}

\noindent with $\eta_k$ the number of vertices with degree $k$, and $\overline{\Delta}_{i-1}=max(\Delta_{i-1})$, and $max(\Delta_{0})=-1$. This changes equation~\ref{criterio} to
